Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-08-19 paper/ms.tex
Add MNRAS TeX template

%\documentstyle[amssymb,12pt,draft,epsf,palatino]{nature-pvd}
\documentclass[twoside]{natureprintstyle}
\bibliographystyle{naturemag}

\usepackage{aas_macros}

\usepackage{epsfig,caption}
\usepackage{color}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{longtable}
% \usepackage{amssymb}
\usepackage{rotating,xcolor}
\usepackage{hyperref}
% \usepackage{tgbonum}
\usepackage[symbol]{footmisc}
\usepackage{placeins}
\usepackage{amsmath}

% Adds line numbers, just for submission stage.
\usepackage[switch]{lineno}
\linenumbers

% Defines footer format for the date.
\usepackage{datetime}

\newdateformat{footerfmt}{%
\THEDAY\,\,\monthname[\THEMONTH]\,\,\THEYEAR}

% Adds current date to the footer.
\usepackage{fancyhdr}
% \fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[RO]{\textsc{\footerfmt\today}}
\fancyfoot[LE]{\textsc{\footerfmt\today}}
\pagestyle{fancy}

\DeclareMathOperator\arcsinh{arcsinh}

\usepackage{fontspec}
\setmainfont{texgyrepagella}[
  Extension = .otf,
  UprightFont = *-regular,
  BoldFont = *-bold,
  ItalicFont = *-italic,
  BoldItalicFont = *-bolditalic,
]

\usepackage[misc]{ifsym}
\usepackage{xcolor}
\usepackage[labelfont=bf]{caption}
\DeclareCaptionLabelSeparator{lsep}{ | } 
\captionsetup{labelsep=lsep}

\newcommand{\RCR}{\ensuremath{R_{\textrm{CR}}}}
\newcommand{\Rot}{\ensuremath{\mathcal{R}}}

\newcommand{\Nbody}{$N$-body}
\newcommand{\SMUGGLE}{SMUGGLE}
\newcommand{\AREPO}{\textsc{AREPO}}
\newcommand{\Msun}{\ensuremath{M_{\odot}}}

\title{Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down}

\author{Angus~Beane$^{1*}$, Lars~Hernquist$^1$, Elena~D'Onghia$^{2,3}$,
Federico~Marinacci$^{4}$, Charlie~Conroy$^{1}$, Jia~Qi$^{5}$,
Laura~V.~Sales$^{6}$, Paul~Torrey$^{5}$ and Mark~Vogelsberger$^{7}$
}

\begin{document}

% Enforces same format on first page as on subsequent pages.
\makeatletter
\let\ps@plain\ps@fancy
\makeatother

\maketitle

{\let\thefootnote\relax\footnote{

\begin{affiliations}
% 1
\item Center for Astrophysics $|$ Harvard \& Smithsonian,  Cambridge, MA, USA.

% 2
\item Department of Physics, University of Wisconsin-Madison, Madison, WI, USA.

% 3
\item Department of Astronomy, University of Wisconsin-Madison, Madison, WI, USA.

% 4
\item Department of Physics \& Astronomy `Augusto Righi', University of Bologna, Bologna, Italy.

% 5
\item Department of Astronomy, University of Florida, Gainesville, FL, USA.

% 6
\item Department of Physics \& Astronomy, University of California, Riverside, CA, USA.

% 7
\item Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA.

$^{*}$ \texttt{\mbox{angus.beane@cfa.harvard.edu}}

\end{affiliations}
}}

% \renewcommand{\thefootnote}{\fnsymbol{footnote}}

\vspace{-3.5mm}
\begin{abstract}

  Elongated bar-like features are ubiquitous in galaxies, occurring at the
  centers of approximately two-thirds of spiral
  disks\cite{2000AJ....119..536E, 2007ApJ...657..790M}.  Due to gravitational
  interactions between the bar and the other components of galaxies, it is
  expected that angular momentum and matter will redistribute over long (Gyr)
  timescales in barred galaxies \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work ignoring the gas
  phase of galaxies has conclusively demonstrated that bars should slow their
  rotation over time due to their interaction with dark matter halos
  \cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
  2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
  2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
  2009ApJ...697..293D}. We have performed a simulation of a Milky Way-like
  galactic disk hosting a strong bar which includes a state-of-the-art model
  of the interstellar medium and a live dark matter halo. In this simulation
  the bar pattern does not slow down over time, and instead remains at a
  stable, constant rate of rotation. This behavior has been observed in
  previous simulations using more simplified models for the interstellar gas
  but it has remained unexplained\cite{1993AA...268...65F,
  2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark
  matter halo act in concert to stabilize the bar pattern speed and prevent the
  bar from slowing down or speeding up. We find that in a Milky Way-like disk,
  a gas fraction of only about \boldmath$5\%$ is necessary for this mechanism
  to operate. This result naturally explains why nearly all observed bars
  rotate rapidly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
  2019MNRAS.482.1733G} and is especially relevant for our understanding of how
  the Milky Way arrived at its present state.

\end{abstract}

\vspace{1cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/mockHST.pdf}
\caption{\textbf{Synthetic Hubble Space Telescope images of our simulations with
and without the interstellar medium.} The upper panels show an \Nbody{} only
simulation while the lower panels show a simulation which includes the SMUGGLE
model for the interstellar medium. Each panel is $20\,\textrm{kpc}$ to a side.
Columns show different points in time, separated by about $1\,\textrm{Gyr}$.
We can see that in the \Nbody{} run, the bar grows in length and strength. In
the SMUGGLE run, the bar remains at approximately the same length and strength
over the course of the simulation. The \Nbody{} model is identical to the
GALAKOS model, discussed in the text. Details on the production of these
synthetic images is given in the Methods section.}\label{fig:overview}
\end{figure*}

\begin{figure}[h!]%
\centering
\includegraphics[width=9cm]{fig/ps_Rbar_torque.pdf}
\caption{\textbf{Properties of the bar in our \boldmath$N$-body and SMUGGLE
simulations of a barred galaxy.} The \textit{upper panel} shows the evolution
of the pattern speed. As expected, the bar in the \Nbody{} run slows down due
to interactions between the bar and the dark matter halo. However, the bar in
the SMUGGLE run does not slow down and instead remains at a constant pattern
speed. The \textit{middle panel} shows the evolution of the bar length. In the
\Nbody{} case, the bar lengthens. This occurs because as the pattern speed
drops, bar-like orbits at larger radii are possible. Stars are captured on
these orbits, lengthening the bar. This process does not occur in the SMUGGLE
cases since the bar pattern speed is not decreasing, and therefore the bar
length remains constant. The \textit{lower panel} shows the torque on the bar
by different components. The solid lines show the torque exerted by the halo
in both the \Nbody{} and SMUGGLE cases. The dashed line shows the torque
exerted by the gas phase in the SMUGGLE run (there is no gas in the \Nbody{}
run). Details on how these properties are calculated is given in the Methods
section.}\label{fig:prop}
\end{figure}

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/halo_wake.pdf}
\caption{\textbf{The wake excited in the dark matter halo.} The dark matter
halo wake is shown in the \Nbody{} case (\textit{middle panel}) and SMUGGLE case
(\textit{right panel}) after $2.6\,\textrm{Gyr}$ of evolution. The
\textit{middle} and \textit{right panels} show a surface density projection in
the $x$-$y$ plane of the dark matter halo after an axisymmetric average has
been subtracted. The solid line indicates the direction of the bar while the
dashed line indicates the direction of the halo wake (both measured by taking
the second Fourier component within a sphere of all material within a radius
of $4\,\textrm{kpc}$). The \textit{left panel} shows the time evolution of
the angle difference between the bar and the halo wake, as measured from the
second Fourier component. After the first Gyr, the angle difference in the
SMUGGLE case is smaller than in the \Nbody{} case by about a factor of two,
reflecting how the dark matter halo in the SMUGGLE case is unable to exert as
negative a torque on the bar as in the \Nbody{} case.}\label{fig:wake}
\end{figure*}

\begin{figure}[t!]%
\centering
\includegraphics[width=9cm]{fig/ps_late_start.pdf}
\caption{\textbf{Pattern speed evolution with a lower initial pattern speed.} We
tested the evolution of our system when gas is added to the \Nbody{} run at a
later time, but with all other simulation parameters kept the same. The setup
is therefore identical to our previous runs just with a bar that is larger,
stronger, and with a lower pattern speed. We find that the pattern speed
evolution is very similar to our fiducial case, except that the bar retains
its original pattern speed. This indicates a mechanism which keeps the bar at
its formation pattern speed, and that there is not a particular pattern speed
which the system tends to.}
\label{fig:snap700}
\end{figure}

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum\cite{1972MNRAS.157....1L}. The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W}. Stellar disks in isolation are prone to developing bars,
\cite{1971ApJ...168..343H} but the presence of a strong spherical potential
(e.g., from a stellar bulge or dark matter halo) acts to stabilize the disk
against bar formation\cite{1973ApJ...186..467O, 1976AJ.....81...30H}. Many
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on a bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in a dark matter halo in a manner
akin to dynamical friction. This wake lags the bar and thus exerts a negative
torque on it. For a review of bar dynamics, see ref.\cite{1993RPPh...56..173S}

The expectation that a dark matter halo acts to slow down a bar is in
conflict with observational estimates of bar pattern speeds. Bar rotation
rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits
as the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators''\cite{2000ApJ...543..704D}. Galaxies with
$\Rot < 1$ are not thought to be stable\cite{1980AA....81..198C}.
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G, 2020MNRAS.491.3655G}.

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar\cite{2011MNRAS.415.1027H}.
Thus, numerical work has shown that the gas phase can have a stronger
influence on a bar than its contribution to the mass of a galaxy would
suggest\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}.

We have performed a simulation of a disk galaxy using the finite-volume
gravito-hydrodynamics code \AREPO{}\cite{2010MNRAS.401..791S}. In \AREPO{},
the fluid is discretized as a moving Voronoi mesh. We use the additional
physics in the galaxy formation module Stars and MUltiphase Gas in GaLaxiEs
(SMUGGLE)\cite{2019MNRAS.489.4233M}. SMUGGLE is a comprehensive and
self-consistent galaxy formation model with a wide range of physical
processes, including radiative heating/cooling, star formation, and stellar
feedback. More detailed information on SMUGGLE is given in the Methods
section. Our disk galaxy is a modified version of the GALAKOS
model\cite{2020ApJ...890..117D}, which consists of a stellar disk, stellar
bulge, and dark matter halo. After $\sim2.5\,\textrm{Gyr}$ of evolution, the
GALAKOS disk forms a bar consistent with the Milky Way bar in terms of pattern
speed and length
($\sim40\,\textrm{km}/\textrm{s}/\textrm{kpc}$\cite{2022arXiv220606207G} and
$\sim4.5\,\textrm{kpc}$\cite{2015MNRAS.450.4050W}, respectively). We modify
this setup to include a gas phase, the details of which are provided in the
Methods section.

A mock stellar light image using filter profiles from the \textit{Hubble Space Telescope} is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the \Nbody{}
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the \Nbody{} case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the \Nbody{} (blue)
and SMUGGLE (orange) runs. The pattern speed in the \Nbody{} case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The
slowing down of the pattern speed in the \Nbody{} case is consistent with a long
line of numerical research on bars in \Nbody{}
simulations\cite{1992ApJ...400...80H, 2000ApJ...543..704D,
2002MNRAS.330...35A, 2002ApJ...569L..83A, 2003MNRAS.341.1179A,
2003MNRAS.346..251O, 2005MNRAS.363..991H, 2006ApJ...637..214M,
2007MNRAS.375..460W, 2009ApJ...697..293D}. However, in the SMUGGLE case the
pattern speed remains constant. After the first Gyr of evolution, we find that
the pattern speed increases by only $\sim10\%$ over the next
$4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern speed for
the \Nbody{} run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the \Nbody{} case grows over
time while it remains roughly constant in the SMUGGLE case. This is also
consistent with previous numerical work, which found that bars tend to grow as
they slow down and the radius of corotation
increases\cite{2000ApJ...543..704D, 2003MNRAS.341.1179A}.

The bottom panel of Fig.~\ref{fig:prop} shows the torque exerted on the bar by
different components. The solid lines indicate the torque on the bar by the
dark matter halo, whereas the dashed line indicates the torque on the bar by
the gas phase. In the \Nbody{} case, the halo exerts a steady negative torque on
the bar, with an average torque from $1$ to $4\,\textrm{Gyr}$ of $-58.0$ in
units of $10^{10}M_{\odot}\,(\textrm{km}/\textrm{s})^2$. The halo in the
SMUGGLE case exerts a similar negative torque on the bar in the first Gyr of
evolution, but after that the halo exerts a much smaller torque on the bar,
averaging only $-7.8$ in the same units and over the same time interval. The gas in
the SMUGGLE case exerts a steady positive torque averaging $11.7$ over $1\,\textrm{Gyr}$ in the same
units.

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
negative torque on the bar can be understood in terms of the halo wake
mechanism. In the \Nbody{} case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
slowing it down\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}.\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce the wake, arresting the process by which the halo can slow the bar
down. We term this process ``exhaustion.''

We can test this interpretation by measuring the angle offset between the halo
wake and the bar. If the wake and the bar are aligned (i.e., there is no angle
offset), then the wake cannot exert a negative torque on the bar. This angle
is plotted in the left panel of Fig.~\ref{fig:wake}, which shows that the
angle offset is larger in the \Nbody{} case than in the SMUGGLE case by about a
factor of two. The center and right panels of Fig.~\ref{fig:wake} show the halo
wake with respect to the location of the bar in the \Nbody{} (center) and SMUGGLE
(right) cases.

The presence of the gas can arrest the process by which additional material in
the dark matter halo can contribute to a wake. However, this does not explain
why the pattern speed in the SMUGGLE case is nearly constant over several Gyr.
Naively, it would be a coincidence that the bar pattern speed remains constant
in the SMUGGLE case, resulting from a chance cancellation of the halo and gas
torques. However, a constant pattern speed in the presence of gas has been
observed in a few simulations of barred galaxies with
gas\cite{1993AA...268...65F, 2010ApJ...719.1470V}. Previous work has argued
this is due to the bar torquing gas inwards, but no explanation has been given
for why it might remain constant.

We propose that an equilibrium mechanism is responsible for the pattern speed
remaining approximately constant. In this scenario, residual negative torque
from the dark matter halo balances out the positive torque from the gas phase.
It has been shown when an analytic bar is forced to rotate at a constant
pattern speed for a few Gyr, the halo exerts almost no torque on the
bar\cite{2022MNRAS.513..768C}. We instead see over a several Gyr time span
that the dark matter halo in our simulation is still able to support some
negative torque.

We argue that the following occurs. First, the bar is not able to slow down
quickly enough due to the positive torque of the infalling gas. This causes
the resonant halo phase space at a particular pattern speed, $\Omega_{p,0}$,
to become exhausted and no longer able to support a negative torque. Second,
the gas is still exerting a positive torque on the bar, and therefore the
pattern speed will again increase. Since at higher pattern speeds the halo has
not yet been totally exhausted, the halo will once again be able to exert a
negative torque on the bar. The pattern speed will then settle at a new value
slightly higher than $\Omega_{p,0}$ where the gas and halo torques cancel.
Over time, the pattern speed should slowly increase as the halo becomes
progressively exhausted.

A clear prediction of this proposed mechanism is that the constant pattern
speed a particular galaxy will end up is somewhat arbitrary. In the real
universe for an isolated galaxy it would be the formation pattern speed of the
bar while in our simulation it is the pattern speed of the bar when gas is
added to the system. We tested this by adding gas to the system at a later
time when the bar has further grown and slowed down with time. In our
particular test, we added the gas at a time when the pattern speed is
$\sim30\,\textrm{km}/\textrm{s}/\textrm{kpc}$. As shown in
Fig.~\ref{fig:snap700}, we find that the pattern speed evolution is very
similar between the two cases (orange and red lines). If anything, the system
with a lower pattern speed seems to speed up more, which is consistent with
our picture since the stronger bar should experience a larger torque from the
gas as it is more efficient at driving gas inflows. We also show in the
Methods that more slowly rotating bars at fixed bar strength are more
efficient at driving gas inflows as well. Nonetheless, when the initial pattern
speed is lower (red line), the addition of gas does not cause the pattern
speed to quickly return to the higher value of our fiducial simulation (orange
line).

We performed two additional tests to explore how our findings depend on the
assumed model of the interstellar medium and the amount of gas present in the
system. First, we performed a simulation of the same disk but with a simpler
model of the interstellar medium\cite{2003MNRAS.339..289S}, closer to standard
methods used in cosmological simulations of galaxy formation. We find that the
pattern speed evolution is nearly the same in this case, and so conclude that
our result is not sensitive to the details of the model for the interstellar
medium. The details of this experiment are given in the Methods section.

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find for our disk, our proposed mechanism
operates as long as the gas contributes about $5\%$ of the total mass of the
disk. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard \Nbody{} simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies with varying gas fractions.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk are given in the Methods
section.

We note that ref.\cite{2020MNRAS.491.3655G} found that the rotation parameter
\Rot{} positively correlates with gas fraction, such that galaxies with higher
gas fractions are rotating more slowly. However, it is not obvious this is in
tension with our result since the gas fraction of galaxies correlates with
other galactic properties\cite{2009ARAA..47..159B}. Furthermore, the
measurement of pattern speeds is a delicate process still prone to large error
bars.

The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R}. Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C}. Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence\cite{2019MNRAS.490.4740B}. And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

Barred galaxies in cosmological simulations of galaxy formation continue to be
in conflict with observations by producing bars which rotate too
slowly\cite{2017MNRAS.469.1054A, 2019MNRAS.483.2721P, 2021AA...650L..16F}.
However, the pattern speeds of bars in both cosmological simulations and the
real universe can be affected by environmental processes not included in our
simulation -- e.g., satellite infall\cite{2011Natur.477..301P},
non-sphericity\cite{2013MNRAS.429.1949A} or rotation\cite{2013MNRAS.434.1287S,
2014ApJ...783L..18L, 2018MNRAS.476.1331C, 2019MNRAS.488.5788C} in the dark
matter halo, or perhaps even the gaseous circumgalactic medium. Naturally,
extending our present work to account for such effects is a crucial next step
in understanding the formation and evolution of galactic bars.

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which have been gas-poor for
extended periods of time to be rotating very slowly. We therefore predict that
observations which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one example of a galaxy which is known to be a slow rotator -- the
low surface brightness galaxy UGC 628\cite{2009AA...499L..25C}. This galaxy
has been studied in detail by ref.\cite{2016MNRAS.463.1751C}, who note that it
indeed has a low gas fraction for galaxies of its type. We predict a general
trend that bars in gas-rich spiral galaxies should rotate quickly while bars
in gas-poor spiral galaxies should rotate slowly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{ref}

\begin{addendum}

>>>>

% region preamble
% mnras_template.tex 
%
% LaTeX template for creating an MNRAS paper
%
% v3.0 released 14 May 2015
% (version numbers match those of mnras.cls)
%
% Copyright (C) Royal Astronomical Society 2015
% Authors:
% Keith T. Smith (Royal Astronomical Society)

% Change log
%
% v3.0 May 2015
%    Renamed to match the new package name
%    Version number matches mnras.cls
%    A few minor tweaks to wording
% v1.0 September 2013
%    Beta testing only - never publicly released
%    First version: a simple (ish) template for creating an MNRAS paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,usenatbib]{mnras}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
% Depending on your LaTeX fonts installation, you might get better results with one of these:
%\usepackage{mathptmx}
%\usepackage{txfonts}

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}

% Allow "Thomas van Noord" and "Simon de Laguarde" and alike to be sorted by "N" and "L" etc. in the bibliography.
% Write the name in the bibliography as "\VAN{Noord}{Van}{van} Noord, Thomas"
\DeclareRobustCommand{\VAN}[3]{#2}
\let\VANthebibliography\thebibliography
\def\thebibliography{\DeclareRobustCommand{\VAN}[3]{##3}\VANthebibliography}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{newtxtext,newtxmath}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared

% endregion

%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Stellar Bars and Gas]{Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down}

\author[A. Beane et al.]{Angus Beane,$^{1}$\thanks{E-mail: angus.beane@cfa.harvard.edu (AB)}
Lars Hernquist,$^{1}$
Elena~D'Onghia,$^{2,3}$
Federico~Marinacci,$^{4}$
Charlie Conroy,$^{1}$
Jia~Qi,$^{5}$\newauthor
Laura~V.~Sales,$^{6}$
Paul~Torrey,$^{5}$
Mark~Vogelsberger$^{7}$
\\
% List of institutions
$^{1}$Center for Astrophysics $|$ Harvard \& Smithsonian,  Cambridge, MA, USA\\
$^{2}$Department of Physics, University of Wisconsin-Madison, Madison, WI, USA\\
$^{3}$Department of Astronomy, University of Wisconsin-Madison, Madison, WI, USA\\
$^{4}$Department of Physics \& Astronomy `Augusto Righi', University of Bologna, Bologna, Italy\\
$^{5}$Department of Astronomy, University of Florida, Gainesville, FL, USA\\
$^{6}$Department of Physics \& Astronomy, University of California, Riverside, CA, USA\\
$^{7}$Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA\\
}

% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

% Enter the current year, for the copyright statements etc.
\pubyear{2015}

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
	Elongated bar-like features are ubiquitous in galaxies, occurring at the
	centers of approximately two-thirds of spiral
	disks \citep{2000AJ....119..536E, 2007ApJ...657..790M}.  Due to gravitational
	interactions between the bar and the other components of galaxies, it is
	expected that angular momentum and matter will redistribute over long (Gyr)
	timescales in barred galaxies \citep{1972MNRAS.157....1L,
	1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work ignoring the gas
	phase of galaxies has conclusively demonstrated that bars should slow their
	rotation over time due to their interaction with dark matter halos
	\cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
	2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
	2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
	2009ApJ...697..293D}. We have performed a simulation of a Milky Way-like
	galactic disk hosting a strong bar which includes a state-of-the-art model
	of the interstellar medium and a live dark matter halo. In this simulation
	the bar pattern does not slow down over time, and instead remains at a
	stable, constant rate of rotation. This behavior has been observed in
	previous simulations using more simplified models for the interstellar gas
	but it has remained unexplained\cite{1993AA...268...65F,
	2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark
	matter halo act in concert to stabilize the bar pattern speed and prevent the
	bar from slowing down or speeding up. We find that in a Milky Way-like disk,
	a gas fraction of only about \boldmath$5\%$ is necessary for this mechanism
	to operate. This result naturally explains why nearly all observed bars
	rotate rapidly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
	2019MNRAS.482.1733G} and is especially relevant for our understanding of how
	the Milky Way arrived at its present state.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
keyword1 -- keyword2 -- keyword3
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%

\section{Introduction}

\section{Methods, Observations, Simulations etc.}

% \subsection{Figures and tables}

% Figures and tables should be placed at logical positions in the text. Don't
% worry about the exact layout, which will be handled by the publishers.

% Figures are referred to as e.g. Fig.~\ref{fig:example_figure}, and tables as
% e.g. Table~\ref{tab:example_table}.

% % Example figure
% \begin{figure}
% 	% To include a figure from a file named example.*
% 	% Allowable file formats are eps or ps if compiling using latex
% 	% or pdf, png, jpg if compiling using pdflatex
% 	\includegraphics[width=\columnwidth]{example}
%     \caption{This is an example figure. Captions appear below each figure.
% 	Give enough detail for the reader to understand what they're looking at,
% 	but leave detailed discussion to the main body of the text.}
%     \label{fig:example_figure}
% \end{figure}

% % Example table
% \begin{table}
% 	\centering
% 	\caption{This is an example table. Captions appear above each table.
% 	Remember to define the quantities, symbols and units used.}
% 	\label{tab:example_table}
% 	\begin{tabular}{lccr} % four columns, alignment for each
% 		\hline
% 		A & B & C & D\\
% 		\hline
% 		1 & 2 & 3 & 4\\
% 		2 & 4 & 6 & 8\\
% 		3 & 5 & 7 & 9\\
% 		\hline
% 	\end{tabular}
% \end{table}

\section{Conclusions}

The last numbered section should briefly summarise what has been done, and describe
the final conclusions which the authors draw from their work.

\section*{Acknowledgements}

The Acknowledgements section is not numbered. Here you can thank helpful
colleagues, acknowledge funding agencies, telescopes and facilities used etc.
Try to keep it short.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Data Availability}

The inclusion of a Data Availability Statement is a requirement for articles published in MNRAS. Data Availability Statements provide a standardised format for readers to understand the availability of data underlying the research results described in the article. The statement may refer to original data generated in the course of the study or to third-party data analysed in the article. The statement should describe and provide means of access, where possible, by linking to the data or providing the required accession numbers for the relevant databases or DOIs.

%%%%%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%%%%

% The best way to enter references is to use BibTeX:

\bibliographystyle{mnras}
\bibliography{ref} % if your bibtex file is called example.bib

% Alternatively you could enter them by hand, like this:
% This method is tedious and prone to error if you have lots of references
%\begin{thebibliography}{99}
%\bibitem[\protect\citeauthoryear{Author}{2012}]{Author2012}
%Author A.~N., 2013, Journal of Improbable Astronomy, 1, 1
%\bibitem[\protect\citeauthoryear{Others}{2013}]{Others2013}
%Others S., 2012, Journal of Interesting Stuff, 17, 198
%\end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Some extra material}

If you want to present additional material which would interrupt the flow of the main paper,
it can be placed in an Appendix which appears after the list of references.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

% End of mnras_template.tex
